Tree growth, mortality, and climatic stress in west-coast states, USA

### 2025-11-03 Manuscript.Rmd
Latest version of manuscript, added CWD analysis to supplemental materials.

@@

@@

@@

fig.cwdmap_cap <- fig_nums(name = "fig_cwdmap", 
                        caption = "Fuzzed FIA plot locations displayed with initial visit climatic water deficit (CWD) values (A) and CWD change between visits (C) for the U.S. states, north to south, of Washington, Oregon, and California (see inset map B for the location of these states within the USA).")

fig.domains <- fig_nums(name = "fig_domains",
                        caption = "Decadal (A) growth and (B) mortality domain estimates for *Calocedrus decurrens*.  The six domains are defined as FIA plots that fall [A]bove or [B]elow the CWD change threshold with [L]ow, [M]edium, or [H]igh initial CWD values (see text for details). E.g., 'AL' is Above CWD change threshold, Low initial CWD value. The distribution of plot CWD change and initial values (C) is presented as well, with the blue horizontal line at a change in CWD of 5 mm representing the change threshold and the green vertical lines representing the quantile boundaries between Low, Medium, and High initial CWD values." )

fig.diff_grow <- fig_nums(name = "fig_diffgrow", 
                          caption = paste0("Differences in mean growth rates (cm\u00B2/decade) between domains by species.  Domain abbreviations are described in the legend for ",
                                           f.ref("fig_domains"), 
                                           ". See each column legend for the key of domain differences.  Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled. Species above the dashed line are coniferous; those below are deciduous. Empirical MC permutation p-values are provided for filled-circle contrasts, with \u2013\u2013 corresponding to p-values of > 0.05, while one to four asterisks correspond to p > 0.01, > 0.001, > 0.0001 and < 0.0001, respectively."))

fig.diff_mort <- fig_nums(name = "fig_diffmort", 
                          caption = paste0("Differences in mean decadal mortality rates between domains by species.  Domain abbreviations are described in the legend for ",
                                           f.ref("fig_domains"), 
                                           ". See each column legend for the key of domain differences.  Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled. Species above the dashed line are coniferous; those below are deciduous.  Empirical MC permutation p-values are provided for filled-circle contrasts, with \u2013\u2013 corresponding to p-values of > 0.05 while one to four asterisks correspond to p > 0.01, > 0.001, > 0.0001 and < 0.0001, respectively. "))

tab.summ_grow_mort <- tab_nums(name = "tbl_summ_grow_mort",
                               caption = paste0("Summary of number of comparisons with 95% CIs that do not overlap with zero ([T]otal, [P]ositive-, and [N]egative-valued estimates) for Figures ", fig_nums("fig_diffgrow", display = "num"), " and ", fig_nums("fig_diffmort", display = "num"), " by analysis (T)ype (Growth or Mortality) and type of (C)omparison.  The column Domains lists the domains compared; All is the sum of values within a Comparison. See text for other domain labels.  'n' is the total number of differences examined within a Comparison."))

@@

There  are signs that tree species in western North America are in regional decline.  Johnston et al. (2025) document a 29% 10-year mortality rate of 300+ yr old trees in Eastern Oregon.  High-severity fires in California in 2020 and 2021 killed 13%-19% of giant sequoia (Sequoiadendron gianteum) with >1.07-m bole diameters (Stephenson and Brigham 2021, Shive et al. 2021); recruitment in these burns was depressed (Soderberg et al. 2022).   Five of eight common species in western North America exhibit declines in relative density  as measured by relative stand density (Stanke et al. 2021).  Shriver et al. (2022) found that four of five juniper and piñon pine species were declining in portions of their range, comparing mortality to recruitment rates.

Climate change as manifest through increases or intensification of drought and aridity can affect tree mortality directly through drought and indirectly by increasing wildfire frequency and severity and promoting insect and disease outbreak.  Hammond et al. (2022) found globally that the intensity of drought and heat events are directly associated with widespread forest die-off events.  Wildfire frequency and area burned in the western United States has increased, in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt increasingly experienced in the western United States (Abatzoglou and Williams 2016, Westerling 2016, Juang et al. 2022).  There has also been a regional increase in the area of bark beetle outbreaks (Meddens et al. 2012, Hicke et al. 2016).  Many of these climate-influence mortality causes may be attributable to the physical condition of trees following abnormal or prolonged exposure to water stress.

Drought-induced mortality can be driven by carbon starvation and hydraulic failure (see reviews by Adams et al. 2010 and McDowell et al. 2008).  Carbon starvation occurs when isohydric species (e.g., the piñon pine group), or those which mainly rely on stomatal closure to avoid excessive water loss through evapotranspiration (ET) receive insufficient atmospheric carbon dioxide for photosynthesis.  Anisohydric plants, or those who do not strongly regulate water loss through stomatal closure, risk hydraulic failure if sufficient xylem cavitation occurs due to insufficient water uptake needed to meet ET demands (e.g., as seen in Populus tremuloides; Anderegg et al. 2015). Adams et al. (2017) performed a review of drought-induced mortality across 23 species and concluded that hydraulic failure occurred broadly across taxa while non-structural carbohydrate reserves at mortality differed across taxa.  Regardless of mechanism, drought (specifically, water deficit) appears to be an important direct driver of mortality for tree taxa (Hammond et al. 2022. Repetitive citation. Others?).

Drought and physiological stress from drought can indirectly increase mortality as well.  Drought stress can increase tree susceptibility to wildfire, insects, and disease.  Wildfire fire area (Abatzoglou and Williams, 2016; Parks et al. 2025) and number (Westerling 2016) are positively associated with increasing aridity in the western US.   Bark beetle outbreaks in semi-arid areas of the western US may be positively affected by or associated with acute intense drought and water-stressed trees (Negrón 2009, Gaylord et al. 2015) but chronically stressed trees may not provide like bark beetle populations with the resources sufficient to promote population expansion (Huberty and Denno, 2004).  Norway spruce (Picea abies) saplings inoculated with the necrotrophic pathogenic fungi Heterobasidion annosum sensu stricto and exposed to drought conditions were more likely to die than control saplings or saplings exposed to either drought conditions or inoculation, although Scots pine (Pinus sylvestris) seedlings did not show a similar relationship (Gomez-Gallego et al. 2022). Other insects and disease may be promoted or inhibited by drought, depending on the circumstances and due to conflicting factors (Kolb et al. 2016).

@@

We used U.S. National Forest Inventory datasets for our analyses of forestland in California, Oregon, and Washington, USA.  Forestland is defined as an area of 4,050 m$^2$ that is at least 10 percent stocked or potentially stocked with tree species, excluding urban and agricultural land uses (Reams et al., 2005).  Additionally, as our study relied on plot revisit data, we only examined plots that were considered forestland during both visits.

FIA plots consist of four points where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied in a 7.32 m radius circular subplot (total area 672.5 m$^2$) with additional trees sampled in an anulus around each subplot (macroplot) around each subplot, with a total radius of 18 m from the center point and a total area of 1012 m$^2$.  We did not include data from FIA microplots.  Bechtold and Scott (2005) provide plot design details and measurement protocols.

Since 2000, FIA plots have been selected via a spatially balanced probability sample (McRoberts 2005, Reams et al., 2005) with a target plot density of one ground plot every 24 km$^2$. Some regions and ownerships intensify the plot density; we excluded these additional plots to simplify stratification estimation, relying on FIA-defined strata and using the density of FIA Phase 1 sampling points within and across strata to determine strata weights.

Plots in the three states are sampled every 10 years with a target of 1/10 of plots sampled in a panel every year (Thompson 2015). Our analysis depends on examining the fate and growth of trees from two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure.  We obtained our FIA data from the USDA FIA DataMart web portal (https://research.fs.usda.gov/products/dataandtools/fia-datamart; downloaded May 5, 2025).  They included plot data sampled from 2001 to 2023.

@@

We obtained plot-level CWD data from ClimateNA (Wang et al. 2016) for June, July, and August (summer months) between 1980 and 2023 (Bryce: I just averaged the CWD values for these months.  It wasn’t available seasonally, was it? Also, can you provide more info on how CWD is calculated?  I know it is PET-AET, but I’m not sure how to interpret a monthly/seasonal value). Latitude and longitude were based on FIA-fuzzed coordinates (citation?  Bryce – should we tighten this up by running the analysis on non-fuzzed coordinates?).  We found the 20-year mean of summer CWD values for each plot prior to and including the first visit.  We refer this this value as the “initial CWD value”. We then found the mean of summer CWD values between visits and including the second visit year.  The difference between this second value and initial CWD value is the “CWD change value” (maybe ‘difference’ really is a better, less loaded term).  Both the initial CWD and CWD change values are mapped in Figure 1.

## Analysis
Estimation approach: Following Section 4.3.4 in Scott et al. (2005) we obtained point estimates of species growth and mortality using a ratio estimators (Eqn 4.16).  For mortality, the numerator is the sum of stratum-weighted means of trees per plot of a given species that died in plots.  The number of sampled trees in each plot is expanded by the area sampled (subplot or macroplot area) adjusted by the proportion of the plot area available for sampling and the number of years between visits.    The denominator is similar but is calculated as the sum of stratum-weighted means of all trees per plot of a given species.

The numerator of species-specific growth estimates is the sum of stratum-weighted means of the increase in basal area (cm$^2$) per plot while the denominator is the same as was used for the mortality estimate. This produces a point estimate for a species’ mean growth.  Mortality and growth values were standardized to a decadal period.

The analysis finds species growth and mortality estimates for six domains based on CWD values (Figure 2C).   The initial CWD values are divided into three quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High).  These initial CWD value divisions allowed for species-specific assessments.  For all species we defined an arbitrary boundary of 5 mm to define plots with a greater change in CWD (Above [the boundary]) and those with less CWD change (Below).  This was done because species did not differ greatly in CWD change (Table S2).  The category Below includes CWD changes that are negative (i.e., mean CWD was larger prior to the initial visit than between visits).  We also report species mortality and growth estimates for all plots across and within Oregon, Washington, and California irrespective of CWD values.  We limited the number of species examined to those with more than 400 plots.  We report domain estimates for all CWD-based domains with 10 or more plots.

We obtained empirical means and 95% confidence intervals via bootstrap. For each of the 1000 bootstrap iterations we sampled FIA plots with replacement and obtained point estimates for each species' domain.  For each domain we calculated a mean and upper and lower 95% CIs with the 2.5% and 97.5% quantiles from the  sampling iterations.  This approach allows for nonsymmetrical CIs around means.  We estimated contrasts between pairs of CWD domains by subtracting within-iteration domain point estimates.

Our analysis approach results in many comparisons.  We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates (p-values). The MC permutation analysis had 1000 iterations; each iteration relied on the bootstrap confidence interval estimation described above. The domain assignments for each species was sampled without replacement.  The distance of MC permutation contrast CI bounds from zero were compared against test CI bounds. The MC permutation procedure recorded a null significant value if the absolute values of the permutation iteration CI bounds were at least as far from zero as observed bounds.

@@

Our study tracked the fate of `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` species in `r comma(analysis.stats$plot.num)` plots that were visited twice by FIA crews. The two visits for `r pct.remeas[2]`% of plots were 10 years apart (`r pct.remeas[1]`% were 9 years, `r pct.remeas[3]`% were 11 years). Plots were distributed in `r analysis.stats$strat.num` strata.  Of the `r length(sel.spp)` species, `r analysis.stats$n.gym.spp` are gymnosperms.  Mean overall growth rates for species varied between `r min.grow.spp$mean` (*`r min.grow.spp$spp`*) and `r max.grow.spp$mean` (*`r max.grow.spp$spp`*, `r t.ref("tbl_growth")`) cm$^2$/decade.  Decadal mean mortality rates varied between `r min.mort.spp$mean` (*`r min.mort.spp$spp`*) and `r max.mort.spp$mean` (*`r max.mort.spp$spp`*, `r t.ref("tbl_mort")`). Growth and mortality rates differed among states (Tables `r tab_nums("tbl_growth", display = "num")` and `r tab_nums("tbl_mort", display = "num")`).

Forested plot 10-year differences in CWD do not visually coincide with initial CWD values (Figure `r fig_nums("fig_cwdmap", display = "num")`); however, `r analysis.stats$cwd.gt0.pct`% of plots increased in their CWD values to some degree during that time period. Tree species exhibited a variety of initial-visit CWD ranges, with means ranging from `r min.mean.init` (*`r min.mean.init.spp`*) to `r max.mean.init` (*`r max.mean.init.spp`*; Table S1).  The CWD difference tended to be comparatively small, with a maximum 95th percentile value of `r max.q95.diff` (*`r max.q95.diff.spp`*, Table S2). The distribution of individual species plot domain values for CWD and change in CWD can be explored on the dashboard (https://tinyurl.com/FIAgrowmort).

The confidence intervals for 10-year mortality and growth rate domains did not necessarily follow the same patterns nor did all confidence intervals for a species' mortality or growth estimates overlap (e.g., Figures `r fig_nums("fig_domains", display = "num")`A and `r fig_nums("fig_domains", display = "num")`B). The relationship between initial CWD values and plot-level change in CWD were not reliably linear (e.g., Figure `r fig_nums("fig_domains", display = "num")`C) and varied by species (see dashboard).

We examined differences in 10-year growth estimates (Figure `r fig_nums("fig_diffgrow", display = "num")`) and mortality estimates (Figure `r fig_nums("fig_diffmort", display = "num")`) among domains.  The column "Above vs. Below Threshold" compares the domains above the CWD change threshold with those below it (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns BL, BM, and BH from AL, AM, AH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A). Negative estimates for the "Above vs. Below Threshold" in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`  indicate that the 10-year growth or mortality rate was greater for a species in plots below the change estimates, respectively.  `r t.ref("tbl_summ_grow_mort")` summarizes the number of domain differences in each of the columns in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` where the 95% CI does not include zero as well as the number of comparisons per figure column.

Growth domain comparisons (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to be produce substantially more "significant" negative vs. positive estimates, and all three columns of Figure `r fig_nums("fig_diffgrow", display = "num")` had similar numbers of "significant" comparisons. Growth may have been susceptible (negatively or positively) to a change in CWD and differed by initial CWD values, and the responses are species-specific. *Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains.  Except for *Alnus rubra*, these results predominantly occurred in both the Below domain (2) or the Above and Below domains (5).  *Thuja plicata* had higher growth in the Above and Below Medium domains than either High or Low domains. *Abies concolor*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Arbutus menziesii*, and *Quercus kelloggii* had greater growth in the more water stressed High domains than the Medium and/or Low domains. Except for *Larix occidentalis*, these results predominantly occurred in both the Above domain (3) or the Above and Below domains (3). Other species exhibited no 'significant' findings or displayed a mixture of results.

>>>>

# Correlation/fits for Initial CWD vs Delta CWD:
CWD.rels <- read_rds(paste0(RESULTS.OTHER, "CWD_deltaCWD_relationship.rds"))
CWD.lin.cor <- CWD.rels$lm1$r.squared^0.5 
CWD.dAIC <- round(CWD.rels$aic.cwd$AIC[1] - CWD.rels$aic.cwd$AIC[2], 0)

@@

grow.tot.sig <- grow.AvB.sig + grow.A.sig + grow.B.sig

@@

mort.tot.sig <- mort.AvB.sig + mort.A.sig + mort.B.sig

@@

fig.cwdmap_cap <- fig_nums(name = "fig_cwdmap", 
                        caption = "Fuzzed FIA plot locations displayed with initial visit CWD values (A) and change in CWD between visits (C) for the U.S. states, north to south, of Washington, Oregon, and California (see inset map B for the location of these states within the USA).")

fig.domains <- fig_nums(name = "fig_domains",
                        caption = "Decadal (A) growth and (B) mortality domain estimates for *Calocedrus decurrens*.  The six domains are defined as FIA plots that fall [A]bove or [B]elow the CWD change threshold with [L]ow, [M]edium, or [H]igh initial CWD values (see text for details). E.g., 'AL' is Above CWD change threshold, Low initial CWD value. The distribution of plot CWD change and initial values (C) is presented as well, with the blue horizontal line indicating the CWD change threshold and the orange vertical lines representing the quantile boundaries between Low, Medium, and High initial CWD values." )

fig.diff_grow <- fig_nums(name = "fig_diffgrow", 
                          caption = paste0("Differences in mean growth rates (cm\u00B2/decade) between domains by species.  Domain abbreviations are described in the legend for Figure ",
                                           f.ref("fig_domains"), 
                                           ". See each column legend for the key of domain differences.  Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled. Species above the dashed line are coniferous; those below are deciduous. Empirical MC permutation p-values are provided for filled-circle contrasts, with \u2013\u2013 corresponding to p-values of > 0.05, while one to four asterisks correspond to p > 0.01, > 0.001, > 0.0001 and < 0.0001, respectively."))

fig.diff_mort <- fig_nums(name = "fig_diffmort", 
                          caption = paste0("Differences in mean decadal mortality rates between domains by species.  Domain abbreviations are described in the legend for Figure ",
                                           f.ref("fig_domains"), 
                                           ". See each column legend for the key of domain differences.  Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled. Species above the dashed line are coniferous; those below are deciduous.  Empirical MC permutation p-values are provided for filled-circle contrasts, with \u2013\u2013 corresponding to p-values of > 0.05 while one to four asterisks correspond to p > 0.01, > 0.001, > 0.0001 and < 0.0001, respectively. "))

tab.summ_grow_mort <- tab_nums(name = "tbl_summ_grow_mort",
                               caption = paste0("Summary of number of contrasts with 95% CIs that do not overlap with zero ([T]otal, [P]ositive-, and [N]egative-valued estimates) for Figures ", fig_nums("fig_diffgrow", display = "num"), " and ", fig_nums("fig_diffmort", display = "num"), " by analysis (T)ype (Growth or Mortality) and type of (C)omparison.  The column Domains lists the domains compared; All is the sum of values within a Comparison. See text for other domain labels.  'n' is the total number of differences examined within a Comparison."))

@@

There  are signs that tree species in western North America are in regional decline.  Johnston et al. (2025) document a 29% 10-year mortality rate of 300+ yr old trees in Eastern Oregon.  Peret et al. (2022) estimated a decadal range-wide mortality rate of 25% for Subalpine fir (*Picea engelmannii*).  High-severity fires in California in 2020 and 2021 killed 13%-19% of giant sequoia (*Sequoiadendron gianteum*) with >1.07-m bole diameters (Stephenson and Brigham 2021, Shive et al. 2021); recruitment in these burns was depressed (Soderberg et al. 2022).   Five of eight common species in western North America exhibit declines as measured by relative stand density (Stanke et al. 2021).  Shriver et al. (2022) found that four of five juniper (*Juniperus* spp.) and piñon pine (*Pinus* spp.) species were declining in portions of their range, comparing mortality to recruitment rates.

Climate change as manifest through increases or intensification in drought and aridity can affect tree mortality directly through physiological damage and indirectly by increasing wildfire frequency and severity and by promoting insect and disease outbreaks.  Hammond et al. (2022) found globally that the intensity of drought and heat events are directly associated with widespread forest die-off events.  Wildfire frequency and area burned in the western United States has increased, in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt (Abatzoglou and Williams 2016, Westerling 2016, Juang et al. 2022).  There has also been a regional increase in the area of bark beetle (subfamily *Scolytinae* spp.) outbreaks (Meddens et al. 2012, Hicke et al. 2016).  Many of these indirect climate-influenced mortality causes may be attributable to the physical condition of trees following abnormal or prolonged exposure to water stress.

Drought-induced mortality can be driven by carbon starvation and hydraulic failure (see reviews by Adams et al. 2010 and McDowell et al. 2008).  Carbon starvation occurs when isohydric species (e.g., piñon pine, *Pinus edulus*; Breshears et al. 2009), or those which mainly rely on stomatal closure to avoid excessive water loss through evapotranspiration (ET), receive insufficient atmospheric carbon dioxide for photosynthesis.  Anisohydric plants, or those who do not strongly regulate water loss through stomatal closure, risk hydraulic failure if sufficient xylem cavitation occurs due to insufficient water uptake needed to meet ET demands (e.g., as seen in *Populus tremuloides*; Anderegg et al. 2015). Adams et al. (2017) performed a review of drought-induced mortality across 26 species and concluded that hydraulic failure occurred broadly across taxa while non-structural carbohydrate reserves at mortality differed across taxa.  Regardless of mechanism, drought (specifically, water deficit) appears to be an important direct driver of mortality for tree taxa (Hammond et al. 2022).

Drought stress can increase tree susceptibility to wildfire, insects, and disease.  Wildfire area (Abatzoglou and Williams, 2016; Parks et al. 2025) and number (Westerling 2016) are positively associated with increasing aridity in the western US.   Bark beetle outbreaks in semi-arid areas of the western US may be positively affected by or associated with acute intense drought and water-stressed trees (Negrón 2009, Gaylord et al. 2015) but chronically stressed trees may not provide bark beetle populations with the resources sufficient to promote population expansion (Huberty and Denno, 2004).  Norway spruce (*Picea abies*) saplings inoculated with the necrotrophic pathogenic fungi *Heterobasidion annosum* sensu stricto and exposed to drought conditions were more likely to die than control saplings or saplings exposed to either drought conditions or inoculation, although Scots pine (*Pinus sylvestris*) seedlings did not show a similar relationship (Gomez-Gallego et al. 2022). Other insects and disease may be promoted or inhibited by drought, depending on the circumstances and due to conflicting factors (Kolb et al. 2016).

One regionally useful metric of water stress is Climatic Water Deficit (CWD; Stephenson 1998) which appears to be a reasonable covariate for examining tree growth and mortality.  Stephenson (1990) defines CWD as evaporative demand not met by available water, or the difference between plant water demand (actual evapotranspiration, AET), and plant water supply (potential evapotranspiration, PET); these variables are driven by temperature and precipitation, respectively.  CWD is geographically readily available from downscaled PRISM data (Daly et al. 2008, Wang et al. 2016).  The Anderegg  et al. (2015) tree mortality study relied on CWD to capture drought occurrence.  Das et al. (2013) found low-elevation California tree mortality best modeled using CWD.  Restaino et al. (2016) found decreased Douglass fir growth with an increase in CWD.

Data from the U.S. National Forest Inventory, collected by the US Forest Service Forest Inventory and Analysis Program (FIA), is well suited for evaluating regional species mortality and growth.   Since 2001 the FIA program has gathered a variety of forestry data at plots along the west coast of the USA using a spatially balanced probability sample (McRoberts 2005, Reams et al. 2005).  The sample allows for a relatively unbaised design-based inference of population characteristics across a large geographical region.  Data from plot revisits and individual tree data permit direct estimation of change features such as mortality and growth estimation.

Monleon and Linz (2015) evaluated the evidence of tree species range shifts by subtracting the design-based mean estimates of annual FIA plot temperatures for seedlings from those of trees. FIA plot revisit data were mostly incomplete at the time of their analysis.  Groom and Monleon (2023) used the revisit data to compare plot climate variable values (vapor pressure deficit [VPD], temperature, and precipitation) to estimate shifts in plot occupancy and density.  Both studies used a single climate variable value associated with each plot for estimation; e.g., Groom and Monleon (2023) examined shifts in species density by estimating the difference in VPD means between plots with more trees of a given species and plots with fewer trees.  They documented, however, that the initial climate variable values could (and often did) change between plot visits, with species often exhibiting positive or negative correlations between the initial visit and between-visits change value of the climate variable.  They identified that their design-based analysis could potentially improve by including both initial visit and change values for the climate variable.

The purpose of this analysis is to obtain design-based estimates of tree species’ growth and mortality across the western U.S. states of California, Oregon, and Washington as they relate to plot-level initial-visit water stress (Climatic Water Deficit, initial CWD) and the change in water stress (delta CWD) over time.  We approached this bivariate analysis by dividing species’ plot data into domains based on initial and delta CWD values, finding growth and mortality estimates for each domain, and estimating domain contrasts to establish differences between domains.  Our secondary goal includes providing decadal growth and mortality estimates for our analysis species overall and by state to establish species status and to inform our CWD analysis findings.

@@

Our analysis relies on FIA data for forestland in California, Oregon, and Washington, USA.  Forestland is defined as an area of 4,050 m$^2$ that is at least 10 percent stocked or potentially stocked with tree species, excluding urban and agricultural land uses (Reams et al., 2005).  We constrained our analysis to only include remeasurements on areas considered forestland during both visits.

FIA plots consist of four points where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied in a 7.32 m radius circular subplot (total area 672.5 m$^2$) with additional trees sampled in an anulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m$^2$.  We did not include data from FIA microplots.  Bechtold and Scott (2005) provide plot design details and measurement protocols.

Since 2000, FIA plots have been selected via spatially-balanced probability sampling (McRoberts 2005, Reams et al., 2005) with a target plot density of one ground plot every 24 km$^2$. Some regions and ownerships intensify the plot density; we excluded these additional plots to simplify stratification estimation, relying on FIA-defined strata and using the density of FIA Phase 1 sampling points within and across strata to determine strata weights.

Plots in the three states are sampled every 10 years with a target of 1/10th of plots sampled in a panel every year (Thompson 2015). Our analysis depends on examining the fate and growth of trees from two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure.  We obtained our FIA data from the USDA FIA DataMart web portal (https://research.fs.usda.gov/products/dataandtools/fia-datamart; downloaded May 5, 2025).  The database included plot data sampled from 2001 to 2023.

@@

We obtained plot-level CWD data from the model TerraClimate (Abatzoglou et al. 2018) for June, July, and August (summer months) between 1980 and 2023. Latitude and longitude were based on FIA-fuzzed coordinates.  We found the 20-year mean of summer CWD values for each plot prior to and including the first visit year.  We refer this this value as the “initial CWD value”. We then found the mean of summer CWD values between visits and including the second visit year.  The difference between this second value and initial CWD value is the “delta CWD value”.  Both the initial CWD and delta CWD values for FIA plots are mapped in Figure 1.

## Analysis
Following Section 4.3.4 in Scott et al. (2005) we obtained point estimates of species growth and mortality using ratio estimators (Eqn 4.16).  The numerator for species-specific mortality is the sum of stratum-weighted means of the number of trees that died between visits per plot.  The number of sampled trees in each plot is expanded by the area sampled (subplot or macroplot area) adjusted by the proportion of the plot area available for sampling and the number of years between visits.    The denominator for mortality is similar but is calculated as the sum of stratum-weighted means of all trees per plot of a given species.

The numerator of species-specific growth estimates is the sum of stratum-weighted means of the increase in basal area (cm$^2$) between visits per plot while the denominator is the same as was used for the mortality estimate. This produces a point estimate for a species’ mean growth.  Mortality and growth values were standardized to a decadal time period.

The analysis finds species growth and mortality estimates for six domains based on CWD values (Figure 2C).   The initial CWD values are divided into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High).  Delta CWD plot values for all species were divided into these above a 5mm CWD change threshold (Above) and those below the threshold (Below). The 5mm threshold was selected because it is approximately the mean delta CWD value across species and because species' delta CWD value distributions did not differ greatly (Table S2).  The category Below includes negative delta CWD values (i.e., mean CWD was larger prior to the initial visit than between visits).  The domain designations AL, AM, and AH correspond to Above Low, Above Medium, and Above High. The Below domains are similarly labeled (BL, BM, BH). We also report species mortality and growth estimates for all plots across and within Oregon, Washington, and California irrespective of CWD values.  We limited the number of species examined to those with more than 400 occupied plots across the three states.  We calculated domain estimates for all CWD-based domains with 10 or more plots.

We examined contrasts in 10-year growth estimates (Figure `r fig_nums("fig_diffgrow", display = "num")`) and mortality estimates (Figure `r fig_nums("fig_diffmort", display = "num")`) among domains.  The figure column "Above vs. Below Threshold" displays the mean point estimate and 95% CI for the contrast between corresponding Above and Below domains (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns BL, BM, and BH from AL, AM, AH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A). Negative estimates for the "Above vs. Below Threshold" in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`  indicate that the 10-year growth or mortality rate was greater for a species in plots below the change estimates.  `r t.ref("tbl_summ_grow_mort")` summarizes the number of domain differences in each of the columns in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` where the 95% CI does not include zero (termed "significant" findings) as well as the number of contrasts per figure column.

We obtained empirical means and 95% confidence intervals via bootstrap. For each of the 1000 bootstrap iterations we sampled FIA plots with replacement and obtained point estimates for each species' domain.  For each domain we calculated a mean, upper, and lower 95% CI with the 2.5% and 97.5% quantiles from the  sampling iterations.  This approach allows for nonsymmetrical CIs around means.  We estimated contrasts between pairs of CWD domains by subtracting within-iteration domain point estimates.

Our analysis approach results in many comparisons.  We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates (p-values) for "significant" contrasts. The MC permutation analysis had 1000 iterations; each iteration estimated each species' bootstrap confidence interval as described above. The plot domain assignments for each species were sampled without replacement.  The distance of MC permutation contrast CI bounds from zero were compared against test CI bounds. The MC permutation procedure recorded a null significant value if the absolute values of the permutation iteration CI bounds were at least as far from zero as observed bounds.

## Dashboard
We developed a data dashboard (https://tinyurl.com/FIAgrowmort) to allow readers to examine, for individual species, a species' domain estimates, the distribution of initial and delta CWD values, and the geographical distribution of plots and plot domains.  All analysis files, manuscript markdowns, dashboard code, and ancillary data sources are available at the GitHub repository https://github.com/jeremygroom/Tree_Growth_Death.

@@

Our study tracked the fate of `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` species in `r comma(analysis.stats$plot.num)` plots that were visited twice by FIA crews. Of the `r length(sel.spp)` species, `r analysis.stats$n.gym.spp` are gymnosperms.  The two visits were 10 years apart for `r pct.remeas[2]`% of plots (`r pct.remeas[1]`% were 9 years, `r pct.remeas[3]`% were 11 years). Plots were distributed in `r analysis.stats$strat.num` strata.

Mean overall growth rates for species varied between `r min.grow.spp$mean` (*`r min.grow.spp$spp`*) and `r max.grow.spp$mean` (*`r max.grow.spp$spp`*, `r t.ref("tbl_growth")`) cm$^2$/decade.  Six species had greater growth (non-overlapping CIs) in California than in Oregon and/or Washington: *Abies concolor*, *Abies grandis*, *Pinus ponderosa*, *Tsuga heterophylla*, *Arbutus menziesii*, and *Notolithocarpus densiflorus*.  Three species had greater growth in Oregon (*Calocedrus decurrens*, *Picea engelmanii*, and *Pseudotsuga menziesii*) and two species had greater growth in Washington (*Abies amablis* and *Pinus monticola*).  Decadal mean mortality rates varied between `r min.mort.spp$mean` (*`r min.mort.spp$spp`*) and `r max.mort.spp$mean` (*`r max.mort.spp$spp`*, `r t.ref("tbl_mort")`). California had four species with greater mortality rates than Oregon and/or Washington: *Abies concolor*, *Pinus ponderosa*, *Acer macrophyllum*, and *Quercus kelloggii*.  Oregon had one species with a greater mortality rate (*Tsuga mertensiana*) and Washington had four (*Abies grandis*, *Larix occidentalis*, *Picea engelmannii*, and *Pseudotsuga menziesii*).

Delta CWD increased at `r analysis.stats$cwd.gt0.pct`% of plots. Initial and delta CWD values were positively linearly associated (correlation = `r CWD.lin.cor`); however, the two variables were better modeled by a 3rd order polynomial ($\Delta AIC = $ `r CWD.dAIC`, Figure S1, see model fits in Tables S1 and S2), and for most species delta CWD neither linearly related to nor monotonically increased with initial CWD values (Figure S2; also, see dashboard).  Tree species exhibited a variety of initial-visit CWD ranges, with means ranging from `r min.mean.init` (*`r min.mean.init.spp`*) to `r max.mean.init` (*`r max.mean.init.spp`*; Table S3).  Delta CWD values tended to be comparatively small, with a maximum 95th percentile value of `r max.q95.diff` (*`r max.q95.diff.spp`*, Table S4).

Individual species growth and mortality domain patterns were not always similar (e.g., Figures `r fig_nums("fig_domains", display = "num")`A and `r fig_nums("fig_domains", display = "num")`B). Growth domain contrasts (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to result in substantially more "significant" negative or positive estimates, and all three columns of Figure `r fig_nums("fig_diffgrow", display = "num")` had over 20 "significant" contrasts. Growth may have been susceptible (negatively or positively) to a change in CWD and differed by initial CWD values, and the responses are species-specific. *Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains.  Except for *Alnus rubra*, these results predominantly occurred in both the Below domain (2 spp.) or the Above and Below domains (5 spp.).  *Thuja plicata* had greater growth in the Above and Below Medium domains than either High or Low domains. *Abies concolor*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Arbutus menziesii*, and *Quercus kelloggii* had greater growth in the more water stressed High domains than the Medium and/or Low domains. Except for *Larix occidentalis*, these results predominantly occurred in both the Above domain (3 spp.) or the Above and Below domains (3 spp.). Other species exhibited no 'significant' findings (4 spp.) or displayed other result patterns (3 spp.).
